Light inextensible strings under tension in the Schwarzschild geometry

### 2019-05-22 string_jpam.tex
units
Branch to address issue #6
put both sets of units side-by-side

\begin{equation}\label{schwarzschild}
ds^2= -dt^2\left(1-1/r\right) +\frac{dr^2}{1-1/r} + r^2\left(d\theta^2 + \sin^2\theta d\phi^2\right),
\end{equation}

\noindent where units in which~$2GM=c=1$ are used, has been known for
over a century~\cite{schwarzschild1916}.  Here I consider the spatial
component~$ds^2= dr^2/\left(1-1/r\right) + r^2d\phi^2$, confined to
equatorial points~($\theta=\pi/2$); we understand~$1\leq r<\infty$
throughout.  A circular string at constant~$r$ will, trivially, have
proper length~$2\pi r$.

Considering first the radial component we may calculate the
proper length $L\left(r_1,r_2\right)$ of a string stretching
from~$r=r_1\geq 1$ to~$r_2\geq r_1$ at constant $\theta,\phi$:

\begin{equation}\label{radial_string_length}
  \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-1/r}}=
  \left.
  \sqrt{r(r-1)} +\frac{1}{2}\log\left(
  \frac{\sqrt{r}+\sqrt{r-1}}{\sqrt{r}-\sqrt{r-1}}\right)
  \right|_{r_1}^{r_2}
  \end{equation}

Observe that integration may be performed with a lower limit
of~$r_1=1$ with no difficulty, even though the radial component of the
Schwarzschild metric approaches infinity there.

Elementary methods show that the proper length $\ell$ of radial strings has
some peculiar features.  Firstly, consider the proper length from
$r=1$ to $r=1+w$ where $w>0, \left|w\right|\ll 1$.  Then

@@

This would indicate that, close to the Schwarzschild radius $r=1$,
large (but finite) amounts of string are required to connect two
points with slightly differing radial coordinates.  Passing to the
case $w\gg 1$ we have an asymptotic expansion that begins

@@

\noindent suggesting that there is ``extra radial space'', of length
over $1.19$ Schwarzschild radii, sensible at large distances from the
event horizon.  Radial strings have proper length longer than the
Schwarzschild radial coordinate difference $w$.

@@

@@

\begin{equation}\label{rdashdash}
  r''\left(\phi\right) =
  (r-1) + \frac{(4r-3)\left(r'\right)^2}{2r\left(r-1\right)},\qquad r>1.
\end{equation}

\noindent 
Solutions to equation~\ref{rdashdash} correspond to a taut string
confined to Flamm's paraboloid~\cite{flamm1916}; see
Figure~\ref{flamm}.  It is interesting to compare
equation~\ref{rdashdash} with the corresponding equation for null
geodesics, which is usually given~\cite{wald} in terms of $u=r^{-1}$
as $u''=3u^2/2-u$ but is equivalent
to~$r''\left(\phi\right)=r-\frac{3}{2}+\frac{2r'}{r}$.

There does not appear to be a simple analytical solution to
equation~\ref{rdashdash} but nevertheless we may make several
observations.  Equation~\ref{rdashdash} is a second-order nonlinear
ordinary differential equation and thus has two constants of
integration.  There are two qualitatively different types of
solutions: those with a tangential point ($r'=0,r>1$), and those
without.  Consider first solutions with such a point: there exists a
pair~$\left(r,\phi\right), r>1$ at which~$r'(\phi)=0$.  This point
must be unique, as~$r''(\phi)>0$, and be the closest approach of the
string to the event horizon.

@@

\begin{equation}\label{roc}
  R = \frac{
    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
  }{
%    \left|
    r^2 + 2\left(r'\right)^2-rr''
%    \right|
  }
  =
  \frac{
    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
  }{
    r-\frac{\left(r'\right)^2}{2\left(r-1\right)}
  }\qquad\mbox{along a taut string}
\end{equation}

\noindent which would give~$R$ as a function of~$\left(r,r'\right)$.
Considering~$r'(\phi)=0$ shows that if a string is tangential at a
point with~$r_0>1$, then it has a radius of curvature of~$r_0^2$ at
that point.  Any circular path, including a circular orbit at~$r=r_0$,
has a radius of curvature of $r_0<r_0^2$, showing that circular orbits
cannot follow a taut string at closest approach.  Also observe that
null geodesics have a radius of curvature of $2r_0^2/3 < r_0^2$,
showing that light cannot follow a taut string at closest approach
either.  The string has a point of inflection
if~$\left(r'\right)^2=2r(r-1)$.

@@

If, conversely, $r'(\phi)\neq 0$ whenever~$r>1$, then the string has
only one free end; the other is tangential to the event horizon (in
the sense that ${\displaystyle \lim_{r\longrightarrow 1^+}r'=0}$),
except in the degenerate case of constant~$\phi$ when the string is
radial for~$r>1$.

@@

Here I solve equation~\ref{rdashdash} using the \verb+deSolve+ package
for ordinary differential equations~\cite{soetart2010}.  Non-radial
solutions fall into two distinct classes: those with $r>1$ at all
points, and those which attain $r=1$, in which case the string is
tangential to the event horizon.

We first consider strings for which $r>1$ everywhere (that is,
completely exterior strings).
Figure~\ref{closest_approach_non_self_intersecting} shows a sequence
of non-self-intersecting numerical solutions to
equation~\ref{rdashdash}, rotated so that the tangential point of
closest approach occurs at~$\phi=0$.
Figure~\ref{closest_approach_self_intersecting} shows the
corresponding diagram for self-interesecting strings
(Figure~\ref{light_closest_approach} shows equivalent null geodesics
for comparison).  A bisection technique reveals that if the closest
approach is greater than a critical value of about~$\simeq 1.0761$,
the string does not self-intersect; if the closest approach is less
than this value, then the string crosses itself at~$\phi=\pi/2$.  As
discussed above, if $\left(r'\right)^2=2r(r-1)$, the string has a
point of inflection at which point the radius of curvature becomes
infinite.  Figure~\ref{curvature_switch} shows non-self intersecting
strings rotated so that the point of inflection occurs at~$\phi=0$; we
see that each string has a finite length in which it is convex to the
origin and two semi-infinite stretches in which the curvature is
negative.

There are other ways to parametrize completely exterior strings;
Figure \ref{strings_r_equals_2}, for example, shows strings passing
through~$(2,0)$.  Figure~\ref{light_r_equals_2} shows light paths for
comparison.

The cases where the string attains $r=1$ are tangent to the event
horizon (or, exceptionally, purely radial).
Figure~\ref{fixed_EH_intersection} shows strings tangential to the
event horizon at~$\phi=0$ and Figure~\ref{one_free_end_r_equals_2}
shows tangential strings passing through $(2,0)$.

@@

\begin{figure}[p] % created by "closest_approach2.R"
\centering
\includegraphics[width=170mm]{closest_approach_selfintersecting.pdf}
\caption{Self-intersecting strings shown for $\phi\geq 0$.  Closest
  approach to the event horizon occurs at $\phi=0$ and
  $r\in\left(1,1.0761\right)$; strings are symmetrical about $\phi=0$}
\label{closest_approach_self_intersecting}
\end{figure}

@@

\begin{figure}[p]
\centering
\includegraphics{angle_at_r_equals_2.pdf} % created by "angle_at_r_equals_2.R"
\caption{Strings passing through $(2,0)$}
\label{strings_r_equals_2}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics{light_start_at_r_equals_2.pdf} % created by "angle_start_at_r_equals_2.R"
\caption{Null geodesics passing through $(2,0)$, shown for $0\leq\phi\leq 2\pi$}
\label{light_r_equals_2}
\end{figure}

@@

\begin{figure}[p] % created by "one_free_end_r_equals_2.R"
\centering
\includegraphics[width=170mm]{one_free_end_r_equals_2.pdf}
\caption{Strings tangential to the event horizon passing through $(2,0)$}
\label{one_free_end_r_equals_2}
\end{figure}

>>>>

\begin{equation}\label{schwarzschild}
ds^2= -dt^2\left(1-2GM/r\right) +\frac{dr^2}{1-2GM/r} + r^2\left(d\theta^2 + \sin^2\theta d\phi^2\right),
\end{equation}

\noindent where units in which~$c=1$ are used, has been known for
over a century~\cite{schwarzschild1916}.  Here I consider the spatial
component~$ds^2= dr^2/\left(1-2GM/r\right) + r^2d\phi^2$, confined to
equatorial points~($\theta=\pi/2$); we understand~$2GM\leq r<\infty$
throughout.  A circular string at constant~$r$ will, trivially, have
proper length~$2\pi r$.

Considering first the radial component we may calculate the
proper length $L\left(r_1,r_2\right)$ of a string stretching
from~$r=r_1\geq 2GM$ to~$r_2\geq r_1$ at constant $\theta,\phi$:

\begin{equation}\label{radial_string_length_properunits}
  \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-1/r}}=
  \left.
  \sqrt{r(r-1)} +\frac{1}{2}\log\left(
  \frac{\sqrt{r}+\sqrt{r-1}}{\sqrt{r}-\sqrt{r-1}}\right)
  \right|_{r_1}^{r_2}
  \end{equation}

\begin{equation}\label{radial_string_length}
  \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-2GM/r}}=
  \left.
  \sqrt{r(r-2GM)} +\frac{1}{2}\log\left(
  \frac{\sqrt{r}+\sqrt{r-2GM}}{\sqrt{r}-\sqrt{r-2GM}}\right)
  \right|_{r_1}^{r_2}
  \end{equation}

Observe that integration may be performed with a lower limit
of~$r_1=2GM$ with no difficulty, even though the radial component of the
Schwarzschild metric approaches infinity there.

Elementary methods show that the proper length $\ell$ of radial strings has
some peculiar features.  Firstly, consider the proper length from
$r=2GM$ to $r=2GM+w$ where $w>0, \left|w\right|\ll 1$.  Then

@@

\begin{equation}
  \ell=\int_{r=2GM}^{r=2GM+w}ds
  =2\sqrt{2GM}w^{1/2} + \frac{w^{3/2}}{3\sqrt{2GM}} + O\left({w}^{5/2}\right)
\end{equation}

This would indicate that, close to the Schwarzschild radius $r=2GM$,
large (but finite) amounts of string are required to connect two
points with slightly differing radial coordinates.  Passing to the
case $w\gg 2GM$ we have an asymptotic expansion that begins

@@

\begin{equation}
  \ell = w +  GM\log\left(\frac{w}{2GM}\right) + GM(1+ 2\log 2)+\frac{GM}{4}w^{-1} + O(w^{-2}),
\end{equation}

\noindent suggesting that there is ``extra radial space'', of length
over $1.19\cdot 2GM$, sensible at large distances from the event
horizon.  Radial strings have proper length longer than the
Schwarzschild radial coordinate difference $w$.

@@

\begin{equation}
%  \int_{p_1}^{p_2}\sqrt{\frac{\left(r'\right)^2}{1-2M/r} + r^2}d\phi=
  \int_{p_1}^{p_2}\left(\frac{\left(r'\right)^2}{1-2GM/r} + r^2\right)^\frac{1}{2}d\phi=
  \int_{p_1}^{p_2}F\left(r,r'\right)d\phi
\end{equation}

@@

\begin{equation}\label{rdashdash_properunits}
  r''\left(\phi\right) =
  (r-1) + \frac{(4r-3)\left(r'\right)^2}{2r\left(r-1\right)},\qquad r>1.
\end{equation}
\begin{equation}\label{rdashdash}
  r''\left(\phi\right) =
  (r-2GM) + \frac{(4r-6GM)\left(r'\right)^2}{2r\left(r-2GM\right)},\qquad r>2GM.
\end{equation}

\noindent 
Solutions to equation~\ref{rdashdash} correspond to a taut string
confined to Flamm's paraboloid~\cite{flamm1916}; see
Figure~\ref{flamm}.  It is interesting to compare
equation~\ref{rdashdash} with the corresponding equation for null
geodesics, which is usually given~\cite{wald} in terms of $u=2GM/r$ as
$u''=3u^2/2-u$, but is equivalent
to~$r''\left(\phi\right)=r-3GM+4GMr'/r$.

There does not appear to be a simple analytical solution to
equation~\ref{rdashdash} but nevertheless we may make several
observations.  Equation~\ref{rdashdash} is a second-order nonlinear
ordinary differential equation and thus has two constants of
integration.  There are two qualitatively different types of
solutions: those with a tangential point ($r'=0,r>2GM$), and those
without.  Consider first solutions with such a point: there exists a
pair~$\left(r,\phi\right), r>2GM$ at which~$r'(\phi)=0$.  This point
must be unique, as~$r''(\phi)>0$, and be the closest approach of the
string to the event horizon.

@@

\begin{equation}\label{roc_properunits}
  R = \frac{
    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
  }{
%    \left|
    r^2 + 2\left(r'\right)^2-rr''
%    \right|
  }
  =
  \frac{
    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
  }{
    r-\frac{\left(r'\right)^2}{2\left(r-1\right)}
  }\qquad\mbox{along a taut string}
\end{equation}

\begin{equation}\label{roc}
  R = \frac{
    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
  }{
%    \left|
    r^2 + 2\left(r'\right)^2-rr''
%    \right|
  }
  =
  \frac{
    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
  }{
    r-\frac{\left(r'\right)^2}{2\left(r-2GM\right)}
  }\qquad\mbox{along a taut string}
\end{equation}

\noindent which would give~$R$ as a function of~$\left(r,r'\right)$.
Considering~$r'(\phi)=0$ shows that if a string is tangential at a
point with~$r_0>2GM$, then it has a radius of curvature of~$r_0^2/2GM$
at that point.  Any circular path, including a circular orbit
at~$r=r_0$, has a radius of curvature of $r_0<r_0^2/2GM$, showing that
circular orbits cannot follow a taut string at closest approach.  Also
observe that null geodesics have a radius of curvature of $r_0^2/3GM <
r_0^2/2GM$, showing that light cannot follow a taut string at closest
approach either.  The string has a point of inflection
if~$\left(r'\right)^2=2r(r-2GM)$.

@@

If, conversely, $r'(\phi)\neq 0$ whenever~$r>2GM$, then the string has
only one free end; the other is tangential to the event horizon (in
the sense that ${\displaystyle \lim_{r\longrightarrow 1^+}r'=0}$),
except in the degenerate case of constant~$\phi$ when the string is
radial for~$r>2GM$.

@@

Here I solve equation~\ref{rdashdash} using the \verb+deSolve+ package
for ordinary differential equations~\cite{soetart2010}.  Non-radial
solutions fall into two distinct classes: those with $r>2GM$ at all
points, and those which attain $r=2GM$, in which case the string is
tangential to the event horizon.

We first consider strings for which $r>2GM$ everywhere (that is,
completely exterior strings).
Figure~\ref{closest_approach_non_self_intersecting} shows a sequence
of non-self-intersecting numerical solutions to
equation~\ref{rdashdash}, rotated so that the tangential point of
closest approach occurs at~$\phi=0$.
Figure~\ref{closest_approach_self_intersecting} shows the
corresponding diagram for self-interesecting strings
(Figure~\ref{light_closest_approach} shows equivalent null geodesics
for comparison).  A bisection technique reveals that if the closest
approach is greater than a critical value of about~$\simeq 1.0761\cdot
2GM$, the string does not self-intersect; if the closest approach is
less than this value, then the string crosses itself at~$\phi=\pi/2$.
As discussed above, if $\left(r'\right)^2=2r(r-2GM)$, the string has a
point of inflection at which point the radius of curvature becomes
infinite.  Figure~\ref{curvature_switch} shows non-self intersecting
strings rotated so that the point of inflection occurs at~$\phi=0$; we
see that each string has a finite length in which it is convex to the
origin and two semi-infinite stretches in which the curvature is
negative.

There are other ways to parametrize completely exterior strings;
Figure \ref{strings_r_equals_2}, for example, shows strings passing
through~$(4GM,0)$.  Figure~\ref{light_r_equals_2} shows light paths for
comparison.

The cases where the string attains $r=2GM$ are tangent to the event
horizon (or, exceptionally, purely radial).
Figure~\ref{fixed_EH_intersection} shows strings tangential to the
event horizon at~$\phi=0$ and Figure~\ref{one_free_end_r_equals_2}
shows tangential strings passing through $(4GM,0)$.

@@

\begin{figure}[p] % created by "closest_approach2.R"
\centering
\includegraphics[width=170mm]{closest_approach_selfintersecting.pdf}
\caption{Self-intersecting strings shown for $\phi\geq 0$.  Closest
  approach to the event horizon occurs at $\phi=0$ (3 o'clock) and
  $r/2GM\in\left(1,1.0761\right)$; strings are symmetrical about $\phi=0$
  but for clarity only $\phi\geq 0$ is shown}
\label{closest_approach_self_intersecting}
\end{figure}

@@

\begin{figure}[p]
\centering
\includegraphics{angle_at_r_equals_2.pdf} % created by "angle_at_r_equals_2.R"
\caption{Strings passing through $(4GM,0)$}
\label{strings_r_equals_2}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics{light_start_at_r_equals_2.pdf} % created by "angle_start_at_r_equals_2.R"
\caption{Null geodesics passing through $(4GM,0)$, shown for $0\leq\phi\leq 2\pi$}
\label{light_r_equals_2}
\end{figure}

@@

\begin{figure}[p] % created by "one_free_end_r_equals_2.R"
\centering
\includegraphics[width=170mm]{one_free_end_r_equals_2.pdf}
\caption{Strings tangential to the event horizon passing through $(4GM,0)$}
\label{one_free_end_r_equals_2}
\end{figure}
